---
title: "DS311 - R Lab Assignment"
author: "Kelly Dam"
date: "11/2/2023"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

* In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
* To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
* After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a. Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
# Number of variables
num_vars <- ncol(mtcars)

# Number of observations
num_obs <- nrows(mtcars)

# Answer:
print("There are total of", num_vars, "variables and", num_obs, "observations in this data set.")
```



b. Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
# Summary statistics
summary_stats <- summary(mtcars)

# Displaying summary statistics
print(summary_stats)

# Checking data types
data_types <- sapply(mtcars, class)

# Counting discrete and continuous variables
num_discrete <- sum(data_types == "factor" | data_types == "integer")
num_continuous <- sum(data_types == "numeric")

# Answer:
print("There are "num_discrete," discrete variables and "num_continuous," continuous variables in this data set.")
```



c. Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
# Calculate mean, variance, and standard deviation for mpg
m <- mean(mtcars$mpg)

v <- var(mtcars$mpg)

s <- sd(mtcars$mpg)

# print(paste("The average of Mile Per Gallon from this data set is ",m, " with variance ",v, " and standard deviation",s, "."))
```



d. Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
# Table 1: Average mpg for each cylinder class
avg_mpg_by_cyl <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)

# Table 2: Standard deviation of mpg for each gear class
sd_mpg_by_gear <- aggregate(mpg ~ gear, data = mtcars, FUN = sd)

# Print the tables
print("Table 1: Average mpg for each cylinder class")
print(avg_mpg_by_cyl)

print("Table 2: Standard deviation of mpg for each gear class")
print(sd_mpg_by_gear)

```



e. Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
# Create a crosstab
crosstab <- table(mtcars$cyl, mtcars$gear)

# Print the crosstab
print("Crosstab: Number of observations for each cylinder and gear class combination")
print(crosstab)

# Find the most common car type
most_common_combination <- which.max(crosstab)




print("The most common car type in this data set is car with," most_common_cyl "cylinders and," most_common_gear, "gears. There are total of," num_observations" cars belong to this specification in the data set.")
```


*** 

### Question 2

Use different visualization tools to summarize the data sets in this question.

a. Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
# Load library for visualization
library(ggplot2)

# Create a boxplot 
ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) + 
  geom_boxplot() + 
  labs(title = "Plant Weight in Different Groups",
       x = "Group",
       y = "Plant Weight")
```

Result:

=> Report a paragraph to summarize your findings from the plot!
The boxplot reveals clear differences in the distribution of plant weights among the three groups. It
appears that both Treatment 1 and 2 groups have higher median plant weights compared to the Control group. Additionally, Treatment 2 shows a wider interquartile range, suggesting greater variability in plant weights within that group.

b. Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}
# Create a histogram
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of MPG in mtcars",
       x = "Miles Per Gallon (mpg)",
       y = "Frequency")
       
# Find the most observed mpg class
max_freq_mpg_class <- cut(mtcars$mpg, breaks = seq(min(mtcars$mpg), max(mtcars$mpg) + 10, by 10))
most_observed_class <- levels(max_freq_mpg_class)[which.max(table(max_freq_mpg_class))]
most_observed_class

print("Most of the cars in this data set are in the class of", most_observed_class, "mile per gallon.")
```



c. Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
# Create a pairs plot
pairs(USArrests, diag.panel = NULL, upper.panel = NULL, lower.panel = NULL) +
  geom_point(aes(color = Murder, shape = Assault), alpha = 0.6) +
  labs(title = "Pairs Plot of USArrests Data",
       x = "Murder",
       y = "Assault")

```

Result:

=> Report a paragraph to summarize your findings from the plot!
The pairs plot provides a view of the relationships between different variables in the USArrests dataset.
Specifically, the scatter plot between Murder and Assault reveals a positive correlation, suggesting 
that states with higher murder rates also tend to have assault rates. The color and shape makes it easier
to distinguish between data points. 


***

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder. 

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url='https://www.jaredlander.com/data/housing.csv',
              destfile='data/housing.csv', mode='wb')
housingData <- read.csv('data/housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```

a. Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
# Aggregation tables
# 1. Aggregation by Boro
aggregate_data_boro <- aggregate(Market.Value.per.SqFt ~ Boro, data = housingData, FUN = mean)

# 2. Aggregation by Year.Built
aggregate_data_year <- aggregate(Market.Value.per.SqFt ~ Year.Built, data = housingData, FUN = mean)

# Print the aggregation tables
print("Aggregation by Boro:")
print(aggregate_data_boro)

print("Aggregation by Year.Built:")
print(aggregate_data_year)

```

b. Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# Enter your code here!
# Scatter plot: Market.Value.per.SqFt vs. Year.Built
scatter_plot_year <- ggplot(housingData, aes(x = Year.Built, y = Market.Value.per.SqFt)) +
  geom_point() +
  labs(title = "Scatter Plot: Market Value per SqFt vs. Year Built",
       x = "Year Built",
       y = "Market Value per SqFt")
       
# Scatter plot: Market.Value.per.SqFt vs. Boro
scatter_plot_boro <- ggplot(housingData, aes(x = Boro, y = Market.Value.per.SqFt)) +
  geom_point() +
  labs(title = "Scatter Plot: Market Value per SqFt vs. Borough",
       x = "Borough",
       y = "Market Value per SqFt")
       
# Box plot: Market.Value.per.SqFt by Boro
box_plot_boro <- ggplot(housingData, aes(x = Boro, y = Market.Value.per.SqFt, fill = Boro)) +
  geom_boxplot() +
  labs(title = "Box Plot: Market Value per SqFt by Borough",
       x = "Borough",
       y = "Market Value per SqFt")
       
# Display the plots
print(scatter_plot_year)
print(scatter_plot_boro)
print(box_plot_boro)

```

c. Write a summary about your findings from this exercise.

=> Enter your answer here!
After analyzing the housing dataset and conducting various exploratory data analysis, the summary statistics provide an overview of the central tendency and spread of the market value per square foot which reveals insights into the distribution of housing prices in New York City. The analysis suggests that further investigation into the relationships between variables such as the impact of the year built and borough on housing prices could provide valuable insights into the factors that explain the variations in housing prices in New York City.

